---
title: "Bootstrapping Differential Expression"
description: "How many can I do?"
author: "Gennaro Calendo"
date: "9/22/2023"
categories: [R, bioinformatics]
---

## Das Boot

I have been playing around with the idea of bootstrapping differential 
expression analyses, inspired by [this](https://hbiostat.org/blog/post/badb/) 
excellent blog post from Professor Frank Harrell on how to do bad biomarker 
research. In the section titled, "Difficulties of picking 'winners'" he writes:

>Efron’s bootstrap can be used to fully account for the difficulty of the biomarker selection task. Selection of winners involves computing some statistic for each candidate marker, and sorting features by these strength-of-association measures. The statistic can be a crude unadjusted measure (correlation coefficient or unadjusted odds ratio, for example), or an adjusted measure. For each of a few hundred samples with replacement from the original raw dataset, one repeats the entire analysis afresh for each re-sample. All the biomarker candidates are ranked by the chosen statistic, and bootstrap percentile confidence intervals for these ranks are computed over all re-samples. 0.95 confidence limits for the rank of each candidate marker capture the stability of the ranks.

Since I do a lot of analyses that involve differential expression testing where
samples are done in triplcate, I was curious if I could apply this resampling
strategy to my work to get a better idea of how often my winners are really 
winners given we often have such small groups of samples. 

This blog post is not about performing this bootstrapping workflow (I'll save
that for later). Rather, I want to explore how many unique bootstrap resamples
we expect to generate given triplicate samples and how often we should expect 
any given pattern to by sampled.

## How many possible unique combinations of the data are there?

Since a typical experiment consists of samples done in triplicate the question 
then becomes, how many unique ways of bootstrapping samples are there? The 
reason I care about *unique* resamples is because when estimating differential
expression we are comparing the mean expression between two groups and therefore
a resample consisting of [control1, control1, control2] will give the same 
mean as resampling [control2, control1, control1].

Since I'm no good at math, to examine this question I'll generate a grid of 
all possibilities and count up the unique combinations.

```{r}
library(data.table)

# Create all combinations of the three samples
samples <- c("A", "B", "C")
dt <- setDT(expand.grid(samples, samples, samples))

# Combine into a single string representing the selected samples
dt[, sample := paste0(Var1, Var2, Var3)]

# Count up the number of letters represented in each string
dt[, `:=`(N_A = stringr::str_count(sample, "A"),
          N_B = stringr::str_count(sample, "B"),
          N_C = stringr::str_count(sample, "C"))]

# Count up the unique counts -- total number of rows gives the unique ways of
#  generating bootstraps for triplicates
(nrow(unique(dt[, .(N_A, N_B, N_C)])))
```

As it turns out, this [question](https://stats.stackexchange.com/a/557626) 
has been asked and answered already and the theoretical answer is given by $2n-1\choose{n}$. So ${2(3)-1\choose{3}}={5\choose3}=10$

Just to be sure, let's try again with 4 letters and check against the theoretical
answer

```{r}
samples2 <- c("A", "B", "C", "D")
dt2 <- setDT(expand.grid(samples2, samples2, samples2, samples2))
dt2[, sample := paste0(Var1, Var2, Var3, Var4)]
dt2[, `:=`(N_A = stringr::str_count(sample, "A"),
           N_B = stringr::str_count(sample, "B"),
           N_C = stringr::str_count(sample, "C"),
           N_D = stringr::str_count(sample, "D"))]
(nrow(unique(dt2[, .(N_A, N_B, N_C, N_D)])))
```

```{r}
choose(2*4-1, 4)
```

So if I want to generate bootstrap resamples for a 3x3 experiment there should be
10 x 10 = 100 unique comparisons that I can make. But how often should we expect 
to see any given pattern in a set of triplicates if we perform a bunch of
bootstraps?

## Pattern counts

What is the expected proportion of each pattern in the triplicate experiment if
we are to resample with replacement? We can find this by taking the proportions
of each unique pattern from above.

```{r}
# Create a string from all of the unique ways to count samples
dt[, patterns := paste0(N_A, N_B, N_C)]

# Find the proportion of each of the possible ways to combine samples
sort(table(dt$patterns) / sum(table(dt$patterns)))
```

We can see that some patterns are more likely than others. For example, we are 
just as likely to select 0 As, 0 Bs, and 3 Cs as we are 3 As, 0 Bs, and 0 Cs. 
This is interesting because it suggests that ~22% of our resamples should 
contain the original samples, ~66% should contain one duplicated sample and 
~11% should contain triplicates of single sample.

We should see this if we generate samples and count the occurrences.

```{r}
set.seed(1011001)

bootSamples <- function() {
  # Generate the random string of selected samples
  s <- paste(sample(c("A", "B", "C"), replace = TRUE), collapse="")
  
  # Count the number of times any individual occurs in the string
  data.table(
    N_A = stringr::str_count(s, "A"),
    N_B = stringr::str_count(s, "B"),
    N_C = stringr::str_count(s, "C")
    )
}

# Generate 100 bootstrap resamples
bootstraps <- replicate(1e2, bootSamples(), simplify = FALSE)
bootstraps <- rbindlist(bootstraps)
bootstraps[, patterns := paste0(N_A, N_B, N_C)]

# Count the frequency of the observed patterns
sort(table(bootstraps$patterns) / sum(table(bootstraps$patterns)))
```

Generating 100 samples gets us close to the theoretical values and on average will converge on the theoretical values.

## Thoughts

This is pretty interesting since it suggests that about 5% of the time (0.22 *
0.22 = 0.048) when resampling two groups of triplicate samples I should expect to get back the same results as in the original analysis.
