---
title: Base R Tips & Tricks
author: ''
date: '2022-05-18'
slug: base-r-tips-tricks
categories: []
tags: []
---

I think the [tidyverse](https://www.tidyverse.org/) is great and I love 
[data.table](https://rdatatable.gitlab.io/data.table/) even more **but** as I've
started to develop [my own R packages](https://coriell-research.github.io/coriell/)
I've really tried to cut down on dependencies as much as possible. Understanding
base R has also proven to be very useful when it comes to manipulating data 
and objects from [Bioconductor](https://www.bioconductor.org/) packages. Here, I 
hope to write down some useful base R (and maybe a some `data.table`) tips and
tricks.

## Example Directory

It's commonly the case that data is located in nested directories where the file
name is some generic name and the description of the file is given by the parent
directory like:

```
data/
  sample1/
    data.tsv
    data.log
  sample2/
    data.tsv
    data.log
  sample3/
    data.tsv
    data.log
```

Where each data.tsv file was constructed with:

`df <- data.frame(X = runif(10), Y = runif(10), G = sample(LETTERS, 10))`

We can use these sample files as an example for exploring some base R functionality.

### Listing all files in a nested directory

```{r}
# Create a vector of all .tsv files in the data directory 
files <- list.files(
  path = ".",
  pattern = "*.tsv",
  recursive = TRUE,
  full.names = TRUE
  )
files
```

### Extract names from the files from the filepaths

This handy function below acts like `stringr::str_extract` to extract a regex
matching the pattern from each file path.

```{r}
names(files) <- regmatches(files, regexpr("sample[1-3]", files))
files
```

### Read all files into a single `data.frame`

Specifying the names attribute above will allow the rownames of the rbind'ed 
data.frame to include the sample name.

**See also:**

`vroom::vroom()` or `data.table::rbindlist(lapply(files, fread))` 

```{r}
df <- do.call(rbind, lapply(files, read.delim))

# Show that each sample's data is present
df[c(1:3, 11:13, 21:23), ]
```

### Create a new column in a `data.frame`

Similar to `dplyr::mutate()` or `data.table[, new := fun(d)]` we can use 
`transform()` to add a new column to the `data.frame`

Below, we use `gsub` to find and replace any of the row indexes added by the 
`rbind` call above with "". This `gsub` function is similar to `stringr::str_remove`. 
We then assign these new 'stripped' names to a variable called `sample_name` and 
assign the result back to `df`.

```{r}
df <- transform(df, sample_name = gsub("\\.[0-9]+$", "", rownames(df)))
head(df)
```

### Summarizing data

`table` can give really useful counts per group

```{r}
table(Sample = df$sample_name, Letter = df$G)
```

`aggregate` works like `group_by() %>% summarize()`

For example if we wanted the mean of the X variable for every sample

```{r}
aggregate(X ~ sample_name, data = df, mean)
```

There is also `tapply`.

```{r}
tapply(df$X, df$sample_name, mean)
```

The `ave` function also groups and summarizes. Here, the mean of each factor 
level is returned for each instance of the factor. Like 
`data.table`'s `df[, mean := mean(X), by = sample_name]`

```{r}
ave(df$X, df$sample_name)
```
