---
title: Benchmarking String Joining Functions in R
author: ''
date: '2022-06-02'
slug: benchmarking-string-joining-functions-in-r
categories: []
tags: []
---

I recently had to concatenate tens of millions of strings into a single column of 
strings (thanks Bismark coverage files...) to be used as a unique rownames for a 
combined data.frame of methylation values for multiple samples. I was surprised 
when even my `data.table` code which was something like, 
`dt[, Location := paste(chr, start, sep = "-")]`, was taking minutes. 

So then, what is the fastest way to combine a bunch of strings? 

Since I really care about this in the context of genetic data I'll simulate 
strings from "Chromosomes" and "Start positions" and concatenate them into a 
single "Loci". e.g. "chr1-45678" 

## Load Libraries

```{r message=FALSE}
library(glue)
library(stringi)
library(stringr)
library(data.table)
library(ggplot2)
```

## Create some test strings

Let's make 100 million of these strings to simulate a real test set size. 

I will also pre-create a vector of delimiter values to be used instead of 
having R recycle these values within the function calls themselves.

```{r}
set.seed(1234)

N <- 1e8
chroms <- sample(paste0("chr", as.character(1:22, "X", "Y")), size = N, replace = TRUE)
starts <- as.character(sample.int(1e5, size = N, replace = TRUE))
seps <- rep("-", N)
```

## Candidates

- `paste`: built-in function for concatenating strings
- `paste0`: built-in function for concatenating strings (`paste(..., sep=""`)
- `sprintf`: wrapper around C function `sprintf` for character formatting of strings
- `stringr::str_c`: `stringr` implementation of `paste`
- `stringi::stri_c`: `stringi` function that `stringr::str_c` wraps
- `glue::glue`: String interpolation. Has to be converted to a vector.

Since the test cases are so large I won't use `bench::mark()` but will instead
opt for `system.time()` and call each expression 10 times. This is probably more
reflective of an actual use case.

## Time it!

```{r}
# Define a function for timing each expression
timeIt <- function(expr, times = 10) {
  results <- vector("list", length = times)
  for (i in 1:times) {
    # Only select the User time and the Elapsed Time from system.time results
    results[[i]] <- system.time(eval(expr))[c(1, 3)]
  }
  data.frame(do.call(rbind, results))
}

# Create a list of functions to test
expressions <- list(
  "paste" = quote(paste(chroms, starts, sep = "-")),
  "paste0" = quote(paste0(chroms, seps, starts)),
  "sprintf" = quote(sprintf("%s%s%s", chroms, seps, starts)),
  "stri_join" = quote(stri_join(chroms, seps, starts, ignore_null = TRUE)),
  "str_c" = quote(str_c(chroms, seps, starts)),
  "glue" = quote(as.vector(glue("{chroms}{seps}{starts}")))
  )

# Time the execution of all functions and bind into a single data.table
dt <- rbindlist(lapply(expressions, timeIt), idcol = "expression")
```

## Plot the results

```{r warning=FALSE}
# Cast values long for plotting
dt.m <- melt(
  dt, 
  id.vars = "expression",
  variable.name = "parameter",
  value.name = "time"
  )

# Plot the results
ggplot(dt.m, aes(reorder(expression, time, median), time)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "String Concatentation Times",
       subtitle = "Concatenating (3) 100 million element vectors",
       x = "Expression",
       y = "Time (s)") +
  facet_wrap(~parameter, scales = "free_y") +
  theme_light() +
  theme(
    axis.text.x = element_text(size = 12, color = "black", angle = 45, hjust = 1),
    plot.title = element_text(size = 18, face = "bold"),
    strip.text = element_text(size = 12, color = "black")
    )
```
